### 2022-09-16 main/chapter_searches.tex
write some on likelihood models --- yields and energy

\subsection{Probability, applied}
\begin{figure}[tp]
\centering
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_abacus_likelihood.pdf}
\caption{%
Likelihood: $\mathrm{Poisson}(0\mid x)$ in
(left) $x$ and
(right) $u = \log x$%
}
\end{subfigure}
\\[.5em]
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_abacus_prior.pdf}
\caption{%
Prior: exponential density in
(left) $x$ and
(right) $u = \log x$%
}
\end{subfigure}
\caption[
Likelihoods and prior densities' transformations with coordinates.
]{%
Likelihoods and prior densities' transformations with coordinates.
The $\mathrm{Poisson}(0\mid x)$ likelihood and exponential $e^{-x}\mathrm{d}x$
prior have the same form (left), but the prior density changes to preserve its unit
area relative to $u = \log x$.
On changing to $\log$ coordinates (right), the likelihood slides horizontally,
but the prior density scales to preserve area.
Both maximum likelihood points are equal, but the maximum prior density moves.
\\[.5em]
Particle lifetimes follow exponential distributions and their maximum
densities are at zero time, but it is incorrect to describe their most probable
lifetimes as zero ---
the `most probable' $\log$ lifetime is not $\log 0$, but equals the mean
lifetime.
\\[.5em]
For zero Poisson data, the maximum likelihood rate is zero, but zero might not
be the most probable rate (if, for example, an experiment has known backgrounds).
}
\label{fig:searches_abacus}
\end{figure}
% science snake
Science involves making predictions, testing those predictions against data,
and reporting what is learned from the results.
Various probabilities are involved in this prediction and learning that have
conventional names to communicate what we intend to do with them.
Most important for prediction is a \textbf{likelihood}
\begin{align}
L(a) &= \prob{b}{a} \\
\label{eqn:searches_likelihood}
\intertext{%
that explores the probabilities assigned to a fixed result $b$ from different
contexts $a$, or how well different hypotheses predicted a result.
A likelihood is clearly a probability, but it is not a probability distribution
--- only the left hand argument $b$ forms a normalized distribution.
Alternatively, fixing the context gives a \textbf{prior}%
}
\pi(b) &= \prob{b}{a}
\label{eqn:searches_prior}
\end{align}
for fixed $a$, which is a normalized probability distribution over $b$.

@@

Early variants of $\cls$ derived it from an identity with posterior tail
probabilities in Poisson or Gaussian special cases~\cite{
Helene1983upper,
pdg1988,
read2000modified,
pdg2022ynf
}
with a contrived frequentist interpretation~\cite{zech1988cls}.
For use in general cases, it is more recently described as a
``modified frequentist'',
``approximate confidence in the signal-only hypothesis''~\cite{
read2000modified,
Read2002cls
}.
High-profile usage in Higgs boson searches at the
Large Electron–Positron Collider~\cite{
read1997optimal,
bock1998lower,
etde1998prospects,
junk1999confidence,
lep2000searches,
lep2003search
}
appears to have propelled $\cls$ to prominence;
these early applications used data deficits to exclude their
background-plus signal models with the test statistic $t(n) = -n$, but
disliked how deficits led to small \pvalues\ for the background-only model,
with which they might have confusingly set negative limits on the Higgs boson
signal cross-section.
Such \pvalues\ are illustrated in Figure~\ref{fig:searches_sb_n} (left)
for small data counts;
in contrast, $\cls$ can only favour the background-only model, due to its
negative values that are shown on the right of the same figure.

Although the statement in Section~\ref{sec:searches_fisher} that \pvalue\
ratios are not ``not \emph{necessarily} misleading'',
Figure~\ref{fig:searches_sb_n} shows a bias from the fact that
no data can give a $\cls$ that favours the signal model;
$\log\cls$ is non-positive with the $-n$ test statistic, and indeed non-negative
for the opposite ordering $t(n) = n$.
The $\log$ likelihood ratio on the actual data does not feature this bias:
it favours background-only for deficits and background-plus-signal for
excesses.
Since for Poisson models the $\log$ likelihood ratio is linear in $n$,
\begin{equation}
\log \frac{L(s + b)}{L(b)}
= n\log\frac{s + b}{b} - s
,
\end{equation}
both $n$ and $-n$ are optimal in the Neyman-Pearson sense.

@@

How can one escape this propaganda?
Likelihood ratios on the actual data escape immediately by bypassing the
censor.
Even inequalities need not be biased --- if taken against a predetermined
threshold, say $5$ events, then models can predict whether $n \leq 5$,
and its truth or falsehood in data can favour either side.%
\footnote{%
This context of a fixed inequality is one condition where the
\pvalue\ ratio is not misleading when interpreted as a likelihood ratio.%
}
Finally, a reader can fall back to not trusting the reporter ---
the statements `\atlas\ reports $n \leq 5$' and logical `$n \leq 5$' are not
identical, since one can infer from the former that in fact $n = 5$.
They know what the message says, but not that it states the whole truth.

@@

Instead of $\Lambda$ itself, various modified maximum likelihood ratio test
statistics enjoy standard usage at the LHC~\cite{cern2011procedure};
these are variably defined as piecewise functions of the fitted signal
parameter $s$, and can include likelihoods from the signal-plus-background
model with constrained maximizations which aim to reduce the impact of
unphysical results such as negative Poisson expectations~\cite{
Feldman:1997qc,
Cowan:2010js
},
and can help to better approximate a likelihood ratio ordering.

A great benefit to these modified maximum likelihood ratios is that their
sampling distributions have known approximations in asymptotic (Gaussian)
limits, which assign \pvalues\ through well-behaved analytic
expressions~\cite{Cowan:2010js}.
Without such formulae, the remaining practical approach to estimating
\pvalues\ is to generate Monte Carlo simulations of alternative data.
Such simulations can be computationally expensive, and even hard to define for
``likelihood'' models that blur the line between actual data and
auxiliary model parameters.

\subsection{Like-lihood modelling}
\TODO{
on ``likelihood'' models ---
we are instructed to pretend that these are likelihoods on auxiliary data
TODO cite.
One is free to engage in such doublethink~\cite{orwell1949nnineteen},
% just as one is free to proclaim that $2 + 2 = 5$
but the fact remains that many constraints do not correspond to predictions of
auxiliary data so are not likelihoods: examples include
scale variations,
theoretical cross-section uncertainties,
non-closure uncertainties (which cover differences between estimates).
simulated sample comparisons,
statistical errors on weighted simulations,
extrapolation uncertainties,
and
all uncertainties on supersymmetric signal yields
(for which we have yet no data).
energy based modelling ---
I know of no tangible benefit from pretending that something is a likelihood
when it is not.
}

\TODO{histfactory}
\TODO{minimize energy not max logf}

To model the consequences of these many factors of uncertainty, each factor is
assigned a real-valued parameter, which when varied changed the expected yields
in the histogram bins of the search.
In this way, we form a statistical model which is a function
\begin{equation}
H_{\!f}(\vec \theta) \to (\vec \mu, \log f)
\end{equation}
from the parameters $\theta$ to Poisson expectations $\vec\mu$ of the histogram
bins and a regularization term $\log f$ that is maximized in fitting.

@@

\TODO{uncertainties: systematics, statistical}

\subsection{Significance}

>>>>

\subsection{Probability, applied}
\label{sec:searches_probability_applied}
\begin{figure}[tp]
\centering
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_abacus_likelihood.pdf}
\caption{%
Likelihood: $\mathrm{Poisson}(0\mid x)$ in
(left) $x$ and
(right) $u = \log x$%
}
\end{subfigure}
\\[.5em]
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=\textwidth]{figures/searches_abacus_prior.pdf}
\caption{%
Prior: exponential density in
(left) $x$ and
(right) $u = \log x$%
}
\end{subfigure}
\caption[
Likelihoods and prior densities' transformations with coordinates.
]{%
Likelihoods and prior densities' transformations with coordinates.
The $\mathrm{Poisson}(0\mid x)$ likelihood and exponential $e^{-x}\mathrm{d}x$
prior have the same form (left), but the prior density changes to preserve its unit
area relative to $u = \log x$.
On changing to $\log$ coordinates (right), the likelihood slides horizontally,
but the prior density scales to preserve area.
Both maximum likelihood points are equal, but the maximum prior density moves.
\\[.5em]
Particle lifetimes follow exponential distributions and their maximum
densities are at zero time, but it is incorrect to describe their most probable
lifetimes as zero ---
the `most probable' $\log$ lifetime is not $\log 0$, but equals the mean
lifetime.
\\[.5em]
For zero Poisson data, the maximum likelihood rate is zero, but zero might not
be the most probable rate (if, for example, an experiment has known backgrounds).
}
\label{fig:searches_abacus}
\end{figure}
% science snake
Science involves making predictions, testing those predictions against data,
and reporting what is learned from the results.
Various probabilities are involved in this prediction and learning that have
conventional names to communicate what we intend to do with them.
Most important for prediction is a \textbf{likelihood}
\begin{align}
L(a) &= \prob{b}{a} \\
\label{eqn:searches_likelihood}
\intertext{%
that explores the probabilities assigned to a fixed result $b$ from different
contexts $a$, or how well different hypotheses predicted a result.
A likelihood is clearly a probability, but it is not a probability distribution
--- only the left hand argument $b$ forms a normalized distribution.
Alternatively, fixing the context gives a \textbf{prior}%
}
\pi(b) &= \prob{b}{a}
\label{eqn:searches_prior}
\end{align}
for fixed $a$, which is a normalized probability distribution over $b$.

@@

Early variants of $\cls$ derived it from an identity with posterior tail
probabilities in Poisson or Gaussian special cases~\cite{
Helene1983upper,
pdg1988,
read2000modified,
pdg2022ynf
}
with a contrived frequentist interpretation~\cite{zech1988cls}.
For use in general cases, it is more recently described as a
``modified frequentist'',
``approximate confidence in the signal-only hypothesis''~\cite{
read2000modified,
Read2002cls
}.
High-profile usage in Higgs boson searches at the
Large Electron–Positron Collider~\cite{
read1997optimal,
bock1998lower,
etde1998prospects,
junk1999confidence,
lep2000searches,
lep2003search
}
appears to have propelled $\cls$ to prominence;
these early applications used data deficits to exclude their
background-plus signal models, sometimes with the test statistic $t(n) = -n$,
and disliked how deficits led to small \pvalues\ for the background-only model,
with which they might have confusingly set negative limits on the Higgs boson
signal cross-section.
Such \pvalues\ are illustrated in Figure~\ref{fig:searches_sb_n} (left)
for small data counts;
in contrast, $\cls$ can only favour the background-only model, due to its
negative values that are shown on the right of the same figure.

Despite the statement in Section~\ref{sec:searches_fisher} that \pvalue\
ratios are not ``not \emph{necessarily} misleading'',
Figure~\ref{fig:searches_sb_n} shows a bias from the fact that
no data can give a $\cls$ that favours the signal model;
$\log\cls$ is non-positive with the $-n$ test statistic, and indeed non-negative
for the opposite ordering $t(n) = n$.
The $\log$ likelihood ratio on the actual data does not feature this bias:
it favours background-only for deficits and background-plus-signal for
excesses.
Since for Poisson models the $\log$ likelihood ratio is linear in $n$,
\begin{equation}
\log \frac{L(s + b)}{L(b)}
= n\log\frac{s + b}{b} - s
,
\end{equation}
both $n$ and $-n$ can be optimal in the Neyman-Pearson sense.

@@

How can one escape this propaganda?
Likelihood ratios on the actual data escape immediately by bypassing the
censor.
Even inequalities need not be biased --- if taken against a predetermined
threshold, say $5$ events, then models can predict whether $n \leq 5$,
and its truth or falsehood in data can favour either side.%
\footnote{%
This context of a fixed inequality is one condition where the
\pvalue\ ratio is not misleading when interpreted as a likelihood ratio.%
}
Finally, a reader can fall back to not trusting the reporter ---
the statements `\atlas\ reports $n \leq 5$' and logical `$n \leq 5$' are not
identical, since one can infer from the former that in fact $n = 5$.
One knows what the message says, but not that it states the whole truth.

@@

Instead of $\Lambda$ itself, various modified maximum likelihood ratio test
statistics enjoy standard usage at the LHC~\cite{cern2011procedure};
these are variably defined as piecewise functions of the fitted signal
parameter $s$, and can include likelihoods from the signal-plus-background
model with constrained maximizations which aim to reduce the impact of
unphysical results such as negative Poisson expectations~\cite{
Feldman:1997qc,
Cowan:2010js
},
and can improve (in a context-dependent sense) the resulting ordering.

These modified maximum likelihood ratios benefit from known approximations to
their sampling distributions in asymptotic (Gaussian) limits,
which can assign their \pvalues\ through well-behaved analytic
expressions~\cite{Cowan:2010js}.
Without such formulae, the remaining practical approach to estimating
\pvalues\ is to generate Monte Carlo simulations of alternative data.
Such simulations can be computationally expensive, ill-defined for
``likelihood'' models that blur the distinction between actual data and
auxiliary constraints on the models' parameters.

\subsection{Like-lihood modelling}
To represent the Standard Model and alternatives, LHC data analysis uses
parametric models to describe the models' Poisson expectations of
collision events and how those expectations change in response to their
parameters.
Parameters often correspond to interpretable, physical quantities and when
those parameters are known only with some uncertainty, the models help to
propagate uncertainty into variations of their outputs.

Uncertainties on imperfectly-known yet non-random parameters, however,
conflict with the frequentist denial of prior probabilities.
To avoid this, all modelling is officially framed as not predicting parameters,
but instead constraining them with likelihood functions that use
either real control data or fictional auxiliary data.

This transformation of priors to auxiliary likelihoods is generally possible
--- in a prior density $f(x)\,\mathrm{d}x$ the function $f(x)$ is a likelihood
(strictly, a Radon-Nikodym derivative~\cite{billingsley2008probability})
that constrains the (possibly un-normalized or `improper') prior measure that
is uniform in $x$~\cite{Cowan:2010js}.
But as discussed in Section~\ref{sec:searches_probability_applied}, the shape
of $f(x)$ is entirely determined by the coordinate, and is therefore a free
choice made by the analyst.
By assigning a form to the auxiliary likelihood, therefore, one is really
choosing a \emph{regularizer} that should be indended to induce preferable
behaviours in the parametric model and its `maximum likelihood' statistics.

These regularized models with constraints from fictitious auxiliary data
are usually often ``likelihood'' models~\cite{
cranmer2012histfactory,
baak2015histfitter,
Besjes_2015,
heinrich2021pyhf
}.
To distinguish them from real likelihood functions on actual data, this
thesis uses the alternative spelling \textbf{\heplikelihood}.

Many of our uncertainties do not correspond to predictions of auxiliary data:
examples include
scale variations,
theoretical cross-section uncertainties,
non-closure uncertainties (which cover differences between estimates),
simulated sample comparisons,
statistical errors on weighted simulations,
extrapolation uncertainties,
and
all uncertainties on supersymmetric signal yields
(for which we have yet no data).
We describe these uncertainties by heuristically
combining theoretical modelling, physical intuition, and convention,
and not only by measuring random data.
Believing that they are also likelihoods on auxiliary data is
doublethink~\cite{orwell1949nnineteen}.

Despite their troublesome nomenclature, \heplikelihood\ models are extremely
useful and effective for describing the predictions of our hypotheses and
extracting interpretable results.
A general \heplikelihood\ model is a function $H$ that takes some parameters $x$
and outputs two things:
expected yields $y$ to predict Poisson data,
and an objective value $E$ that should be minimized in a `fit' to choose the
best prediction.
That is,
\begin{equation}
H(x) = [\,y(x), \,E(x)\,]
.
\end{equation}
The minimization objective $E$ is the summed negative $\log$ likelihood
(or regularizer) for all actual and auxiliary data;
we choose the letter $E$ for energy in the analogy to statistical physics where
at low temperature $T = 1/\beta$ a system collapses to its minimuum energy
state due to the Boltzmann factor
\begin{equation}
p(x) \propto e^{-\beta E(x)}
.
\end{equation}
At unit temperature, this can define a propability distribution for the
parametesr relative to an appropraite (prior)
measure~\cite{cranmer2012histfactory, skilling2017david},
but in practice such distributions do not always coincide with intuition or
with the best fit value.

\TODO{fit, expand for error}

\TODO{software tools}

@@

\TODO{uncertainties: systematics, statistical --- maybe susy step?}

\TODO{toy procedure}

\subsection{Gaussian significance}
